Nonlinear Tidal Dissipation in Binary White Dwarfs

### 2019-02-19 proposal/FINESST/Su_FINESST_2019.tex
just don't mask visc, stash proposal changes

    % \everymath{\displaystyle} % biggify limits of inline sums and integrals
    \tikzstyle{circ} % usage: \node[circ, placement] (label) {text};
        = [draw, circle, fill=white, node distance=3cm, minimum height=2em]
    \definecolor{commentgreen}{rgb}{0,0.6,0}
    \lstset{
        basicstyle=\ttfamily\footnotesize,
        frame=single,
        numbers=left,
        showstringspaces=false,
        keywordstyle=\color{blue},
        stringstyle=\color{purple},
        commentstyle=\color{commentgreen},
        morecomment=[l][\color{magenta}]{\#}
    }
    \usepackage[
        backend=bibtex8,
        style=nature]{biblatex}
\defbibenvironment{bibliography}
  {\noindent}
  {\unspace}
  {\printtext[labelnumberwidth]{%
     \printfield{labelprefix}%
     \printfield{labelnumber}}%
   \addspace}
\renewbibmacro*{finentry}{\finentry\addspace}
\bibliography{Su_FINESST_2019}
\AtEveryBibitem{\clearfield{title}}
\begin{document}

@@

\section{Background and Proposed Research}

@@

    \item I will extend my simulations to realistic WD models and equations of
        state such as those in~\cite{brassard1992} as well as those generated by
        MESA (see \autoref{s:3} below), continuing to track the location and
        spatial extent of the dissipation layer as well as any new phenomena. As
        WDs vary widely in composition and effective temperature, studying
        representative WD models is vital to obtaining a robust characterization
        of tidal dissipation.
\end{itemize}

\section{Tidal Heating Thermodynamics}\label{s:3}

@@

As discussed earlier, compact WD binaries exhibit a range of observed and
hypothesized optical transients. Tidal novae\cite{tidal_novae}, underluminous
supernovae\cite{underlum}, and Ca-rich fast transients\cite{carich} are all
hypothesized to arise in WD binary systems. Given that tidal heating can become
a significant contributor to a WD's total energy budget, a realistic model of
tidal dissipation is important to understanding a binary WD's thermodynamic
evolution during inspiral.

In~\cite{tidal_novae} (hereafter FL), the authors use MESA to study the
production of tidal novae in binary WDs. A simple two-zone parameterization
where the tidal heat is deposited throughout the outer zone was used to model
tidal dissipation. They find that cool WDs in sufficiently compact binaries
(orbital period $\lesssim 15$ minutes) may incur a thermonuclear detonation of
the hydrogen envelope. WD binaries with such short orbital periods have been
observed, e.g.\ SDSS J065133+284423 has a period of 12.75 minutes\cite{12min}.
FL fits the observed properties of SDSS J065133+284423 to their tidal heating
models and finds evidence for tidal heating of the secondary WD in SDSS
J065133+284423. From their fits, FL are able to estimate the temperature of the
WD in the absence of tidal heating. As WDs form around a well known temperature
and cool predictably over time, accounting for tidal heating equates to a large
correction to the age of the WD\@. This illustrates the importance of accurately
modeling tidal interactions when inferring physical properties.

@@

With the tidal dissipation models I will develop, I will be able to perform a
simulation similar to that in FL but with a more realistic tidal interaction.
The second aim of my proposal is accordingly: \textbf{I will use my tidal
% "obtained in Section 2"
dissipation profiles to simulate a binary WD undergoing tidal heating and make
comparisons to observational data.} The location of the tidal heating is a key
ingredient in determining whether deposited energy can be efficiently transfered
away or whether the WD experiences sudden detonations. As such, an improved
understanding of the properties of tidal dissipation is important to proper
forecasting of binary WDs' thermodynamic evolution. Moreover, while only a few
sufficiently compact WD binaries were available at the time of writing of FL,
\emph{Gaia} data releases 2 and 3 will provide many more compact WD binaries to
examine for evidence of tidal heating.

Using MESA, I will evolve various WD models undergoing tidal heating. From these
simulations, I will extract the increased temperature of the WDs and make
comparisons to observational data, in particular to WDs in new \emph{Gaia} data
releases. I will also identify the occurrence rate and observational properties
of any predicted optical transients such as tidal novae and attempt to identify
them among existing detected events. Predictions of the occurence rates of such
phenomena are vital to guiding future observations. Finally, comparison of
observational data from known WD binaries to simulations could yield new
insights to the behavior of degenerate matter.

\section{Tidal Heating and LISA}\label{s:4}

As discussed earlier, WD binaries are a primary source of GW radiation for
LISA\@. LISA will attain optimal sensitivity at frequencies
$10^{-4}$--$10^{-1}\;\mathrm{Hz}$\cite{LISA_band}. Exactly in this frequency
range, tidal effects act to synchronize the orbit of the WDs to the binary
orbit and transfer energy from the orbit into the WDs. While the decay
of the binary orbit is still mostly driven by GW radiation, the tidal energy
dissipation rate grows to $\sim10^{-2}$ the GW energy
dissipation\cite{fullerII,fullerIV}. An effect of such a magnitude causes the
phase of the emitted GWs to significantly from the point-mass binary
prediction; the emitted wave may exhibit ``missing cycles'' due to tidal
effects\cite{fullerII}. GW astronomy uses matched filtering, where a library of
template waveforms is matched against instrument data, to identify GW signals.
As such, the accuracy and completeness of the template library is of utmost
importance.

The final aim of my proposal is then: \textbf{I will use my tidal dissipation
model to compute WD binary GW waveforms including tidal dissipation for use in
the LISA detection pipeline}. This aim is much less computationally expensive
than it appears: LISA-band WD binaries can be well described using the Newtonian
dynamics of two co-orbiting point masses\cite{DWD_pointmass}. The resultant GW
emission can then be accurately computed using the weak gravity quadrupole
approximation (see e.g.~\cite{peters,lsst_wd}). Under these two approximations,
the GW waveform can be computed analytically without resorting to numerical
relativity simulations at all, an enormous computational savings. Thus, I will
compute GW waveforms accounting for the additional phase evolution due to tidal
dissipation. I will publish my corrected waveforms for use by LISA and the GW
community.

@@

During the first year of work and first half of the second year, I will complete
my tidal dissipation model. I anticipate the extension of my 2D plane-parallel
work to more complex geometries will be complete within the first half year,
while the extension to realistic WD models will take up to a year. I expect that
these two results together will produce one peer-reviewed publication in
addition to the one currently in preparation.

@@

My work will improve understanding of possible energetic phenomena that can
occur in compact object binaries. My results concerning angular momentum
transfer via IGW breaking will be applicable to astrophysical systems beyond WD
binaries, for instance angular momentum transfer in stars\cite{l_trans_rev}. I
will devise new ways of analyzing astrophysically interesting systems in
\emph{Gaia} and LSST\@. Finally, my GW templates will be important for LISA GW
detection efforts.

@@

>>>>

    % \everymath{\displaystyle} % biggify limits of inline sums and integrals
    \tikzstyle{circ} % usage: \node[circ, placement] (label) {text};
        = [draw, circle, fill=white, node distance=3cm, minimum height=2em]
    \definecolor{commentgreen}{rgb}{0,0.6,0}
    \lstset{
        basicstyle=\ttfamily\footnotesize,
        frame=single,
        numbers=left,
        showstringspaces=false,
        keywordstyle=\color{blue},
        stringstyle=\color{purple},
        commentstyle=\color{commentgreen},
        morecomment=[l][\color{magenta}]{\#}
    }
    \usepackage[
        backend=bibtex8,
        style=nature]{biblatex}
% \defbibenvironment{bibliography}
%   {\noindent}
%   {\unspace}
%   {\printtext[labelnumberwidth]{%
%      \printfield{labelprefix}%
%      \printfield{labelnumber}}%
%    \addspace}
% \renewbibmacro*{finentry}{\finentry\addspace}
\bibliography{Su_FINESST_2019}
\AtEveryBibitem{\clearfield{title}}
\begin{document}

@@

\section{Background and Proposed Research}\label{s:1}

@@

    \item I will extend my simulations to realistic WD models and equations of
        state such as those in~\cite{brassard1992} as well as those generated by
        MESA\cite{MESA} (see \autoref{s:3} below), continuing to track the
        location and spatial extent of the dissipation layer as well as any new
        phenomena. As WDs vary widely in composition and effective temperature,
        studying representative WD models is vital to obtaining a robust
        characterization of tidal dissipation.
\end{itemize}

\section{Tidal Heating and Binary White Dwarf Evolution}\label{s:3}

@@

As discussed earlier, compact WD binaries may exhibit a range of transient
phenomena: tidal novae\cite{tidal_novae}, underluminous
supernovae\cite{underlum}, and Ca-rich fast transients\cite{carich} are all
hypothesized to arise in WD binary systems. Given that tidal heating can become
a significant contributor to a WD's total energy budget, a realistic model of
tidal dissipation is important to understanding the thermal evolution of WDs
during their binary inspiral.

In~\cite{tidal_novae} (hereafter FL), the authors used MESA\cite{MESA} to study
the production of tidal novae in binary WDs. A simple two-zone parameterization
where the tidal heat is deposited throughout the outer zone was used to model
tidal dissipation. It was found that cool WDs in sufficiently compact binaries
(orbital period $\lesssim 15$ minutes) may incur a thermonuclear detonation of
the hydrogen envelope. WD binaries with such short orbital periods have been
observed, e.g.\ SDSS J065133+284423 has a period of 12.75 minutes\cite{12min}.
FL fitted the observed properties of SDSS J065133+284423 to their tidal heating
models and finds evidence for tidal heating of the secondary WD in SDSS
J065133+284423.

@@

With the tidal dissipation models I will develop, I will be able to perform a
binary WD evolution study similar to that in FL but with a realistic tidal
dissipation profile (instead of their parameterized model). The second aim of my
proposal is to \textbf{use my tidal dissipation profiles to simulate a binary WD
undergoing tidal heating and make comparisons to observational data.} The
location of the tidal heating is a key ingredient in determining whether
the deposited tidal energy can be efficiently transfered away or whether the WD
experiences sudden detonations. As such, an improved understanding of the
properties of tidal dissipation is important to proper forecasting of binary
WDs' thermodynamic evolution. Moreover, while only a few sufficiently compact WD
binaries were available at the time of writing of FL, \emph{Gaia} data releases
2 and 3 will provide many more compact WD binaries to examine for observational
manifestations of tidal heating.

Using MESA, I will evolve various WD models undergoing tidal heating. From these
studies, I will extract the increased temperature of the WDs and make
comparisons to observational data, in particular to WDs in new \emph{Gaia} data
releases. I will also identify the occurrence rate and observational properties
of any predicted optical transients such as tidal novae and attempt to identify
them among existing detected events. Predictions of the occurence rates of such
phenomena are vital to guiding future observations. Finally, comparison of
observational data from known WD binaries to my theoretical predictions could
yield new constraints on the interior properties of WDs (e.g.\ crystallization).

\section{Tidal Dissipation and LISA}\label{s:4}

As discussed in \autoref{s:1}, WD binaries are an important source of GW
radiation for LISA\@. LISA will attain optimal sensitivity at frequencies
$10^{-4}$--$10^{-1}\;\mathrm{Hz}$\cite{LISA_band}. Exactly in this frequency
range, tidal effects act to synchronize the spin of the WDs to the binary orbit
and transfer energy from the orbit into the WDs. While the decay of the binary
orbit is still mostly driven by GW radiation, the tidal energy dissipation rate
grows to $\sim10^{-2}$ the GW luminosity\cite{fullerII,fullerIV}. An effect of
such a magnitude causes the phase of the emitted GWs to deviate significantly
from the point-mass binary prediction; the emitted wave may exhibit ``missing
cycles'' due to tidal effects\cite{fullerII}. GW astronomy uses matched
filtering, where a library of template waveforms is matched against instrument
data, to identify GW signals. As such, the accuracy and completeness of the
template library is of utmost importance.

The final aim of my proposal is \textbf{to use my tidal dissipation model to
compute WD binary GW waveforms including tidal dissipation for use in the LISA
detection pipeline}. This aim is much less computationally expensive than it
appears: LISA-band WD binaries can be well described using the Newtonian
dynamics of two co-orbiting point masses\cite{DWD_pointmass}. The resultant GW
emission can then be accurately computed using the weak gravity quadrupole
approximation (see e.g.~\cite{peters,lsst_wd}). Under these two approximations,
the GW waveform can be computed analytically without resorting to numerical
relativity simulations at all. Thus, I will compute GW waveforms accounting for
the additional phase evolution due to tidal dissipation. I will publish my
corrected waveforms for use by LISA and the GW community.

@@

During the first year of work and first half of the second year, I will complete
calculations of tidal dissipation models. I anticipate that the extension of my
2D plane-parallel work to 3D spherical geometries will be complete within the
first half year, while the extension to realistic WD models will take up to a
year. I expect that these two results together will produce two peer-reviewed
publications in addition to the one currently in preparation.

@@

My work will improve understanding of possible energetic phenomena that can
occur in compact object binaries. My results concerning angular momentum
transfer via IGW breaking will be applicable to astrophysical systems beyond WD
binaries, for instance angular momentum transfer in massive
stars\cite{l_trans_rev}. I will devise new ways of analyzing astrophysically
interesting systems in \emph{Gaia} and LSST\@. Finally, my GW templates will be
important for LISA GW detection efforts.

@@

\clearpage
